\propertitle

### 2019-06-19 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-19T14:57:21+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\usepackage[bigdelims,nosymbolsc%,smallerops % probably arXiv doesn't have it
]{newpxmath}
\useosf\linespread{1.083}
%% smaller operators for old version of newpxmath
\makeatletter
\def\re@DeclareMathSymbol#1#2#3#4{%
    \let#1=\undefined
    \DeclareMathSymbol{#1}{#2}{#3}{#4}}
%\re@DeclareMathSymbol{\bigsqcupop}{\mathop}{largesymbols}{"46}
%\re@DeclareMathSymbol{\bigodotop}{\mathop}{largesymbols}{"4A}
\re@DeclareMathSymbol{\bigoplusop}{\mathop}{largesymbols}{"4C}
\re@DeclareMathSymbol{\bigotimesop}{\mathop}{largesymbols}{"4E}
\re@DeclareMathSymbol{\sumop}{\mathop}{largesymbols}{"50}
\re@DeclareMathSymbol{\prodop}{\mathop}{largesymbols}{"51}
\re@DeclareMathSymbol{\bigcupop}{\mathop}{largesymbols}{"53}
\re@DeclareMathSymbol{\bigcapop}{\mathop}{largesymbols}{"54}
%\re@DeclareMathSymbol{\biguplusop}{\mathop}{largesymbols}{"55}
\re@DeclareMathSymbol{\bigwedgeop}{\mathop}{largesymbols}{"56}
\re@DeclareMathSymbol{\bigveeop}{\mathop}{largesymbols}{"57}
%\re@DeclareMathSymbol{\bigcupdotop}{\mathop}{largesymbols}{"DF}
%\re@DeclareMathSymbol{\bigcapplusop}{\mathop}{largesymbolsPXA}{"00}
%\re@DeclareMathSymbol{\bigsqcupplusop}{\mathop}{largesymbolsPXA}{"02}
%\re@DeclareMathSymbol{\bigsqcapplusop}{\mathop}{largesymbolsPXA}{"04}
%\re@DeclareMathSymbol{\bigsqcapop}{\mathop}{largesymbolsPXA}{"06}
\re@DeclareMathSymbol{\bigtimesop}{\mathop}{largesymbolsPXA}{"10}
%\re@DeclareMathSymbol{\coprodop}{\mathop}{largesymbols}{"60}
%\re@DeclareMathSymbol{\varprod}{\mathop}{largesymbolsPXA}{16}
\makeatother
%%
%% With euler font cursive for Greek letters - the [1] means 100% scaling
\DeclareFontFamily{U}{egreek}{\skewchar\font'177}%
\DeclareFontShape{U}{egreek}{m}{n}{<-6>s*[1]eurm5 <6-8>s*[1]eurm7 <8->s*[1]eurm10}{}%
\DeclareFontShape{U}{egreek}{m}{it}{<->s*[1]eurmo10}{}%
\DeclareFontShape{U}{egreek}{b}{n}{<-6>s*[1]eurb5 <6-8>s*[1]eurb7 <8->s*[1]eurb10}{}%
\DeclareFontShape{U}{egreek}{b}{it}{<->s*[1]eurbo10}{}%
\DeclareSymbolFont{egreeki}{U}{egreek}{m}{it}%
\SetSymbolFont{egreeki}{bold}{U}{egreek}{b}{it}% from the amsfonts package
\DeclareSymbolFont{egreekr}{U}{egreek}{m}{n}%
\SetSymbolFont{egreekr}{bold}{U}{egreek}{b}{n}% from the amsfonts package
% Take also \sum, \prod, \coprod symbols from Euler fonts
\DeclareFontFamily{U}{egreekx}{\skewchar\font'177}
\DeclareFontShape{U}{egreekx}{m}{n}{%
       <-7.5>s*[0.9]euex7%
    <7.5-8.5>s*[0.9]euex8%
    <8.5-9.5>s*[0.9]euex9%
    <9.5->s*[0.9]euex10%
}{}
\DeclareSymbolFont{egreekx}{U}{egreekx}{m}{n}
\DeclareMathSymbol{\sumop}{\mathop}{egreekx}{"50}
\DeclareMathSymbol{\prodop}{\mathop}{egreekx}{"51}
\DeclareMathSymbol{\coprodop}{\mathop}{egreekx}{"60}
\makeatletter
\def\sum{\DOTSI\sumop\slimits@}
\def\prod{\DOTSI\prodop\slimits@}
\def\coprod{\DOTSI\coprodop\slimits@}
\makeatother
\input{definegreek.tex}% Greek letters not usually given in LaTeX.

@@

Regarding~\ref{item:maxent_step}, we assume that you're familiar with the
maximum-entropy method. In short, it amounts to a pair of prescriptions:
choose the distribution having maximum Shannon entropy among those
satisfying specific convex constraints, such as fixed averages; and judge
these averages to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First, the distribution
$\px(\yA)$ given by this method is the zeroth-order approximation of four
different distributions for the full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{probability} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the \emph{probability} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximate answer to four
completely different inferences. Note that the four distributions above
numerically differ in higher-order approximations. We discuss their
differences in \sect***. Second, the maximum-entropy method based on the
Shannon entropy implicitly makes some assumptions about the probabilities
for the long-run frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these
assumptions in \sect***.

@@

We can therefore apply the maximum-entropy method to obtain a distribution
for the full network, by constraining its $m$th factorial moment to be
equal to the sample's recorded average of
$\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$, for as many $m$ as we please with
$1\le m \le n$. The constraints on the maximum-entropy distribution
$\px(\yA)$ for the full network are
\begin{equation}
  \label{eq:final_constraint}
  \frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  \equiv
   \sum_{\ya}
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1} \yff_{\ya}
  =
   \sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \px(\yA)
\end{equation}
for any $m$s we wish.

@@

@@

We first calculate the maximum-entropy distribution for the full network by
using six moments. This number already provides as much information as the
full frequency distribution of the sample, and at the same time illustrates
the use of the approach in questions of statistic sufficiency (typically
limited to two or three moments). Figure~\ref{fig:plots_6mom} shows the
resulting densities (distribution/activity-step, that is,
distribution${}\times N$) for full-network sizes $N=n$, $N=1\,000$,
$N=5\,000$, $N=10\,000$\mynote{motivate?}. The case $N=n$ corresponds to
applying the maximum-entropy method at the sample level; it can be observed
that with six moments it reproduces almost exactly the measured frequency
distribution.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{scripts/paper_stensola_6mom_distributionsN_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{scripts/paper_riehle_6mom_distributionsN_prior2.pdf}%
\caption{***}
\label{fig:plots_6mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

The distribution for the full-network is sharper than the measured
frequency distribution for the sample; the sharper the larger $N$ is. Most
remarkably, it has two distinct low-activity modes for the first data set.
But also the second data set presents, upon closer inspection, a small
shoulder on the right of the mode, suggestive of two activity regimes.
These features are clearly not present in the sample or in the
maximum-entropy distribution at the sample level. The application of the
probability calculus to make forecasts about the full network thus reveals
interesting and useful features.

Thereafter we show the results for two moments and for four moments using
$N=10\,000$.

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{scripts/paper2_stensola_twomom_distributions10000_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{scripts/paper_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_6mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-19T19:49:04+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\usepackage[bigdelims,nosymbolsc%,smallerops % probably arXiv doesn't have it
]{newpxmath}
\linespread{1.083}%\useosf
%% smaller operators for old version of newpxmath
\makeatletter
\def\re@DeclareMathSymbol#1#2#3#4{%
    \let#1=\undefined
    \DeclareMathSymbol{#1}{#2}{#3}{#4}}
%\re@DeclareMathSymbol{\bigsqcupop}{\mathop}{largesymbols}{"46}
%\re@DeclareMathSymbol{\bigodotop}{\mathop}{largesymbols}{"4A}
\re@DeclareMathSymbol{\bigoplusop}{\mathop}{largesymbols}{"4C}
\re@DeclareMathSymbol{\bigotimesop}{\mathop}{largesymbols}{"4E}
\re@DeclareMathSymbol{\sumop}{\mathop}{largesymbols}{"50}
\re@DeclareMathSymbol{\prodop}{\mathop}{largesymbols}{"51}
\re@DeclareMathSymbol{\bigcupop}{\mathop}{largesymbols}{"53}
\re@DeclareMathSymbol{\bigcapop}{\mathop}{largesymbols}{"54}
%\re@DeclareMathSymbol{\biguplusop}{\mathop}{largesymbols}{"55}
\re@DeclareMathSymbol{\bigwedgeop}{\mathop}{largesymbols}{"56}
\re@DeclareMathSymbol{\bigveeop}{\mathop}{largesymbols}{"57}
%\re@DeclareMathSymbol{\bigcupdotop}{\mathop}{largesymbols}{"DF}
%\re@DeclareMathSymbol{\bigcapplusop}{\mathop}{largesymbolsPXA}{"00}
%\re@DeclareMathSymbol{\bigsqcupplusop}{\mathop}{largesymbolsPXA}{"02}
%\re@DeclareMathSymbol{\bigsqcapplusop}{\mathop}{largesymbolsPXA}{"04}
%\re@DeclareMathSymbol{\bigsqcapop}{\mathop}{largesymbolsPXA}{"06}
\re@DeclareMathSymbol{\bigtimesop}{\mathop}{largesymbolsPXA}{"10}
%\re@DeclareMathSymbol{\coprodop}{\mathop}{largesymbols}{"60}
%\re@DeclareMathSymbol{\varprod}{\mathop}{largesymbolsPXA}{16}
\makeatother
%%
%% With euler font cursive for Greek letters - the [1] means 100% scaling
\DeclareFontFamily{U}{egreek}{\skewchar\font'177}%
\DeclareFontShape{U}{egreek}{m}{n}{<-6>s*[1]eurm5 <6-8>s*[1]eurm7 <8->s*[1]eurm10}{}%
\DeclareFontShape{U}{egreek}{m}{it}{<->s*[1]eurmo10}{}%
\DeclareFontShape{U}{egreek}{b}{n}{<-6>s*[1]eurb5 <6-8>s*[1]eurb7 <8->s*[1]eurb10}{}%
\DeclareFontShape{U}{egreek}{b}{it}{<->s*[1]eurbo10}{}%
\DeclareSymbolFont{egreeki}{U}{egreek}{m}{it}%
\SetSymbolFont{egreeki}{bold}{U}{egreek}{b}{it}% from the amsfonts package
\DeclareSymbolFont{egreekr}{U}{egreek}{m}{n}%
\SetSymbolFont{egreekr}{bold}{U}{egreek}{b}{n}% from the amsfonts package
% Take also \sum, \prod, \coprod symbols from Euler fonts
\DeclareFontFamily{U}{egreekx}{\skewchar\font'177}
\DeclareFontShape{U}{egreekx}{m}{n}{%
       <-7.5>s*[0.9]euex7%
    <7.5-8.5>s*[0.9]euex8%
    <8.5-9.5>s*[0.9]euex9%
    <9.5->s*[0.9]euex10%
}{}
\DeclareSymbolFont{egreekx}{U}{egreekx}{m}{n}
\DeclareMathSymbol{\sumop}{\mathop}{egreekx}{"50}
\DeclareMathSymbol{\prodop}{\mathop}{egreekx}{"51}
\DeclareMathSymbol{\coprodop}{\mathop}{egreekx}{"60}
\makeatletter
\def\sum{\DOTSI\sumop\slimits@}
\def\prod{\DOTSI\prodop\slimits@}
\def\coprod{\DOTSI\coprodop\slimits@}
\makeatother
\input{definegreek.tex}% Greek letters not usually given in LaTeX.

@@

Regarding~\ref{item:maxent_step}, we assume that you're familiar with the
maximum-entropy method. In short, it amounts to a pair of prescriptions:
choose the distribution having maximum Shannon entropy among those
satisfying specific convex constraints, such as fixed averages; and judge
these averages to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First, the distribution
$\px(\yA)$ given by this method is the zeroth-order approximation \citep[in
the sense of Laplace's
method:][\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four different distributions for the full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{probability} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the \emph{probability} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximate answer to four
completely different inferences. Note that the four distributions above
numerically differ in higher-order approximations. We discuss their
differences in \sect***. Second, the maximum-entropy method based on the
Shannon entropy implicitly makes some assumptions about the probabilities
for the long-run frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these
assumptions in \sect***.

@@

We can therefore apply the maximum-entropy method to obtain a distribution
for the full network, by constraining its $m$th factorial moment to be
equal to the sample's recorded average of
$\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$, for as many $m$ as we please with
$1\le m \le n$. The constraints on the maximum-entropy distribution
$\px(\yA)$ for the full network are
\begin{equation}
  \label{eq:final_constraint}
  \frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  \equiv
   \sum_{\ya}
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1} \yff_{\ya}
  =
   \sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \px(\yA)
\end{equation}
for all $m$ we wish.

@@

For concreteness's sake we'll consider the maximum-entropy distribution
$\px(\yA)$ as \emph{the most probable frequency distribution} for the
full-network activity during the recording; but remember that it is also
the approximation of three other distributions, as discussed in the
previous section.

@@

We first calculate the distribution by using six moments. This number
already provides as much information as the full frequency distribution of
the sample, and at the same time illustrates the use of the approach in
questions of statistic sufficiency (typically limited to two or three
moments). Figure~\ref{fig:plots_6mom} shows the resulting densities
(distribution/activity-step, that is, distribution${}\times N$) for
full-network sizes $N=n$, $N=1\,000$, $N=5\,000$,
$N=10\,000$\mynote{motivate?}. The case $N=n$ corresponds to applying the
maximum-entropy method at the sample level; it can be observed that with
six moments it reproduces almost exactly the measured frequency
distribution.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{scripts/paper_stensola_6mom_distributionsN_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{scripts/paper_riehle_6mom_distributionsN_prior2.pdf}%
\caption{***}
\label{fig:plots_6mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

The distribution for the full-network is sharper than the measured
frequency distribution for the sample; the sharper the larger $N$ is. Most
remarkably, it has two distinct low-activity modes for the first data set.
But also the second data set presents, upon closer inspection, a small
shoulder on the right of the mode, suggestive of two activity regimes. We
frame no hypotheses about the biological cause of these two modes (they
could stem from the presence of different kinds of cells or modules). These
features are clearly not present in the sample or in the maximum-entropy
distribution at the sample level. The application of the probability
calculus to make forecasts about the full network thus reveals interesting
and useful features.

To illustrate how our approach can be applied to studies of sufficient
statistics, \fig~\ref{fig:plots_2_4mom} shows the results for two
constrained moments (equivalent to constraining means and correlations
only) and for four constrained moments, with $N=10\,000$. In either data
set we obtain two very distinct distributions. For the first data set in
particular, four moments lead to a bimodal distribution, whereas two to a
unimodal one, showing that two moments would not be a sufficient
statistics. For comparison, the two distributions obtained applying
maximum-entropy \emph{at the sample level} are shown as an inset in the
plot for the first data set: their differences aren't so glaring as in the
full-network application.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{scripts/paper2_stensola_twomom_distributions10000_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{scripts/paper_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_2_4mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

\section{Marginalization to the sample level}
\label{sec:marginalization}

\section{Assumptions and corrections}
\label{sec:assumptions_formulae}

in depth discussion about underlying assumptions, derivation from
probability calculus, higher-order approximations beyond maxent and
"error bars" around the distributions

\section{Summary and discussion}
\label{sec:summary_discussion}

\clearpage
\hrule
OLD TEXT
\hrule
